Analysis of competitive surfing tournaments with generalized Bradley-Terry

### 2021-08-21 inst/surfing_jsa.tex
tidyup

\section*{Abstract}
Here we apply a generalization of Bradley-Terry model to competitive
surfing, specifically the 2018 World Surf League.  We use
Bradley-Terry likelihoods to assess a number of hypotheses of interest
to the surfing community with formal statistical methods.  We go on to
study the ``Brazilian storm'' phenomenon using reified Bradley-Terry
likelihood functions in two ways: firstly, the supposed Brazilian
preference for point and beach break beaches; and secondly from the
perspective of tournament theory in which we test for competitors
modifying their strategy in the presence of compatriot rivals.

@@

We note in passing that `wct8`, hosted at Lemoore had a different
format from the other venues in the championship, being held in an
artificial wave pool in which every wave was essentially identical (up
to a mirror reflection).  Statistically, there were seven heats of
either 2, 4, 8, or 16 surfers.  We show below how to incorporate the
information present in such observations with the remainder of the
championship using a consistent and intuitive statistical model.

@@

\begin{table}
\begin{verbatim}
wct01 Colapinto Bailey Wright
wct01 Freestone Lau Smith
wct01 Dora Ferreira Slater
wct01 Duru Toledo Ibelli
wct01 Moniz Heazlewood Wilson
...
wct11 Medina Florence
wct11 Colapinto Bourez
wct11 Ferreira Slater
wct11 Medina Colapinto
wct11 Ferreira Medina
\end{verbatim}
\caption{Raw results\label{rawresults}}
\end{table}

@@

We may ask whether the top two officially ranked competitors, Medina
and Ferreira, have different strengths.  To test the null of equal
strength, we maximize the likelihood subject to the null, and compare
with the global maximum likelihood~\cite{hankin2010}.  This gives us a
support difference of about 0.05, well short of Edwards's 2 units of
support criterion.  Alternatively, we could observe that the
$-2\log\left(\frac{\Lambda}{\Lambda_0}\right)\simeq 0.1$ is not in the tail
region of its asymptotic $\chi^2_1$ distribution.  We may thus assert
that there is no strong evidence for Ferreira's actually having a
competitive advantage over Medina.  {\bf Tim: how about comparing
  Medina with, say, Wilson, would that make sense from a surfing
  perspective? Tim can you say something about Wilson's rivalry with
  Medina?}  We may similarly wonder whether Medina is equal in
strength to Wilson; here we reject the null with support difference of
about 2.33, corresponding to a $p$-value of $0.031$.

@@

The statistical method described here can test many different
suggestions drawn from surfing lore {\bf Tim is there a better way to
  describe this?}.  For example, one frequent suggestion {\bf Tim I
  need citations here} is that Brazilian surfers such as Ferreira and
Toledo have surfing skillsets that favour point and beach break
beaches {\bf Tim I need some help describing and defining these here},
those types of beach being readily accessible to these surfers' native
Brazil.  We now assess the null that Brazilians perform equally well
on all types of beaches.  To that end, we create a reified surfer,
``Brazilian beachtype", whose strength helps all Brazilian surfers
when surfing on their (putative) preferred beachtype, analogously to
the ``home ground monster" who helps the home team as per
equation~\ref{home_ground_advantage}.  The null would be that this
reified entity has zero strength.  We fail to reject this null
(differential support = 0.035; $p=0.79$) and conclude that there is no
evidence to suggest that Brazilians' performce is better on their
preferred beach types than one would expect by chance.  We may go
further using the method of support and state that not only is the
maximum likelihood estimate for Brazilian beachtype is very small
(about 0.002), but we may rule out the suggestion that the strength of
this entity exceeds about 0.025 (Figure~\ref{brazilian_beachtype_support}).

@@

{\bf\large Tim I am on thin ice here, I need to talk with you and
  clarify this}

One measure of the rising popularity of surfing is its inclusion in
the 2020 Summer Olympics, and it is natural to ask whether this would
affect the competition in some way.  One suggestion {\bf Tim?} would
be that competitors would become more focussed on beating compatriots
than surfers from other countries, the thinking being that admission
to a country's Olympic squad would be of more importance to a
competitor than their result in wct.  {\bf Tim this is clumsy
  wording}.  We would expect competitors to value their ranking when
restricted to within their country more than their ranking over all of
wct.  If this is the case then we would expect competitors to change
their behaviour if they are in the water with another surfer of the
same nationality.  Specifically, if we have three surfers in the water
of whom two are of one nationality and the third of another, we would
expect the compatriots to adopt a style calculated to beat their
countryman, possibly at the expense of performance relative to the odd
man out.  {\bf Tim this is poor wording we need to rephrase}.  Such
hypotheses are difficult to test with conventional techniques as
competitors's strengths themselves are nuisance
parameters~\cite{basu1977}.

>>>>

\section*{Abstract}
Here we analyse competitive surfing, specifically the 2018 World Surf
League, using formal statistical methods.  We use generalized
Bradley-Terry likelihoods to assess a number of hypotheses of interest
to the surfing community.  Specifically, we quantify the dominance of
the top competitors using likelihood techniques, and go on to study
the ``Brazilian storm'' phenomenon using reified entities in two ways.
Firstly we assess the supposed Brazilian preference for point and
beach break beaches; and secondly we consider tournament results from
the perspective of tournament theory and test for competitors
modifying their strategy in the presence of compatriot rivals.  We
quantify the evidence for these commonly assumed features of
contemporary competitive surfing and suggest further avenues of
research.

@@

We note in passing that Lemoore had a different format from the other
venues in the championship, being held in an artificial wave pool in
which every wave was essentially identical (up to a mirror
reflection).  Statistically, there were seven heats of either 2, 4, 8,
or 16 surfers.  We show below how to incorporate the information
present in such observations with the remainder of the championship
using a consistent and intuitive statistical model.

@@

\begin{table}
\begin{verbatim}
wct01 Colapinto Bailey Wright
wct01 Freestone Lau Smith
wct01 Dora Ferreira Slater
wct01 Duru Toledo Ibelli
wct01 Moniz Heazlewood Wilson
...
wct11 Medina Florence
wct11 Colapinto Bourez
wct11 Ferreira Slater
wct11 Medina Colapinto
wct11 Ferreira Medina
\end{verbatim}
\caption{Extract from observations: raw results\label{rawresults} from
  WCT 2018}
\end{table}

@@

We may ask whether the top two officially ranked competitors, Medina
and Ferreira, have different strengths.  To test the null of equal
strength, we maximize the likelihood subject to the null, and compare
with the global maximum likelihood~\cite{hankin2010}.  This gives us a
support difference of about 0.05, well short of Edwards's 2 units of
support criterion.  Alternatively, we could observe that the
$-2\log\left(\frac{\Lambda}{\Lambda_0}\right)\simeq 0.1$ is not in the
tail region of its asymptotic $\chi^2_1$ distribution.  We may thus
assert that there is no strong evidence for Ferreira's actually having
a competitive advantage over Medina.  We may similarly wonder whether
Medina is equal in strength to Wilson; here we reject the null with
support difference of about 2.33, corresponding to a $p$-value of
$0.031$.

@@

The statistical method described here can test many different
suggestions drawn from surfing lore.  For example, one frequent
suggestion {\bf Tim I need citations here} is that Brazilian surfers
such as Ferreira and Toledo have surfing skillsets that favour point
and beach break beaches {\bf Tim I need some help describing and
  defining these here}, those types of beach being readily accessible
to these surfers' native Brazil.  We now assess the null that
Brazilians perform equally well on all types of beaches.  To that end,
we create a reified surfer, ``Brazilian beachtype", whose strength
helps all Brazilian surfers when surfing on their (putative) preferred
beachtype, analogously to the ``home ground monster" who helps the
home team as per equation~\ref{home_ground_advantage}.  The null would
be that this reified entity has zero strength.  We fail to reject this
null (differential support = 0.035; $p=0.79$) and conclude that there
is no evidence to suggest that Brazilians' performce is better on
their preferred beach types than one would expect by chance.  We may
go further using the method of support and state that not only is the
maximum likelihood estimate for Brazilian beachtype is very small
(about 0.002), but we may rule out the suggestion that the strength of
this entity exceeds about 0.025
(Figure~\ref{brazilian_beachtype_support}).

@@

One measure of the rising popularity of surfing is its inclusion in
the 2020 Summer Olympics~\cite{tulloch2019}, and it is natural to ask
whether this would affect the competition in some way~\cite{ho2021}.
One suggestion would be that competitors would become more focussed on
beating compatriots than surfers from other countries, the thinking
being that admission to a country's Olympic squad would be of more
importance to a competitor than their result in wct.  One might expect
competitors to value their ranking when restricted to within their
country more than their ranking over all of wct.  If this is the case
then we would expect competitors to change their behaviour if they are
in the water with another surfer of the same nationality.
Specifically, if we have three surfers in the water of whom two are of
one nationality and the third of another, we would expect the
compatriots to adopt a style calculated to beat their countryman,
possibly at the expense of performance relative to the odd man out.

Such hypotheses
are difficult to test with conventional techniques as competitors's
strengths themselves are nuisance parameters~\cite{basu1977}.
